Storm Data Analysis - Week 4

### 2021-03-23 Manuscript.Rmd
Switched selection of cases to only include cases with at least one of: injuries, fatalities or damages

mdata <- fdata %>%
    mutate(
        year = as.integer(lubridate::year(begin_time)),
        prop_mult = case_when(
            PROPDMGEXP %in% c(0, 1, 2, 3, 4, 5, 6, 7, 8) ~ 10,
            PROPDMGEXP %in% c('m', 'M') ~ 1000000,
            PROPDMGEXP %in% c('k', 'K') ~ 1000,
            PROPDMGEXP %in% c('b', 'B') ~ 1000000000,
            PROPDMGEXP %in% c('h', 'H') ~ 100,
            PROPDMGEXP == '+' ~ 1,
            # Vector must be double because of limitations in `case_when`, so we
            # can't set a value to NA; use dummy code and convert later.
            TRUE ~ -999999
        ),
        crop_mult = case_when(
            PROPDMGEXP %in% c(0, 1, 2, 3, 4, 5, 6, 7, 8) ~ 10,
            CROPDMGEXP %in% c('m', 'M') ~ 1000000,
            CROPDMGEXP %in% c('k', 'K') ~ 1000,
            CROPDMGEXP %in% c('b', 'B') ~ 1000000000,
            CROPDMGEXP %in% c('h', 'H') ~ 100,
            CROPDMGEXP == '+' ~ 1,
            # Vector must be double because of limitations in `case_when`, so we
            # can't set a value to NA; use dummy code and convert later.
            TRUE ~ -999999
        ),
        # Change dummy value into NA
        across(ends_with("_mult"), ~na_if(., -999999)),
        crop_dmg = CROPDMG * crop_mult,
        prop_dmg = PROPDMG * prop_mult,
        tot_dmg = prop_dmg + crop_dmg
    ) 
```

We can get the number of invalid `EXP` values resulting in `0`-value multipliers using the following code:

```{r invalid-exps}
# Property damage
inv_prop_exp <- filter(mdata, PROPDMG != 0, prop_mult == 0) %>%
    select(PROPDMG, PROPDMGEXP, prop_mult, BGN_DATE) %>%
    nrow()

inv_crop_exp <- filter(mdata, CROPDMG != 0, crop_mult == 0) %>%
    select(CROPDMG, CROPDMGEXP, crop_mult, BGN_DATE) %>%
    nrow()
```

@@

skimr::skim(fdata$INJURIES)
cat("\n\n\n")
skimr::skim(fdata$FATALITIES)

@@

# Make a copy of data and cast as DT
dt_data <- as.data.table(lean_data)
# Define the number of rows that will be in the grouped DT. This is equal to the
# unique values of event, as we'll group by that.
denominator <- uniqueN(dt_data, "event")
# Apply final transformations to DT
sel_data <- dt_data[, .(
                        counter = .N, # Count elements in group
                        # Total monetary damages per group
                        total_dmg = sum(tot_dmg, na.rm = TRUE),
                        # Total fatalities
                        fatalities = sum(FATALITIES, na.rm = TRUE),
                        # Injuries
                        injuries = sum(INJURIES, na.rm = TRUE)
                        ),
                    # Grouping by event type
                    by = event][
                        ,
                        # Compute a "percent rank": rank events by number of
                        # occurrences (desc), then assign a percentile by
                        # dividing rank by denominator (i.e. nrows)
                        # Note double dot syntax to access calling scope
                        `:=`(perc_rank = frankv(counter)/..denominator,
                             # Same for monetary damage
                             dmg_rank = frankv(total_dmg)/..denominator,
                             # And life damage
                             inj_rank = frankv(injuries)/..denominator,
                             fat_rank = frankv(fatalities)/..denominator
                             )
                        ][
                            # Filter for frequency rank >=95th
                            perc_rank >= 0.95
                        ][
                            order(-counter)
                        ]

@@

sel_data %>% ggplot(
    aes(x = fct_reorder(event, injuries, .desc = T), y = inj_rank)
    ) +
    geom_point()

>>>>

temp_data <- fdata %>%
    mutate(
        year = as.integer(lubridate::year(begin_time)),
        prop_mult = case_when(
            PROPDMGEXP %in% c(0, 1, 2, 3, 4, 5, 6, 7, 8) ~ 10,
            PROPDMGEXP %in% c('m', 'M') ~ 1000000,
            PROPDMGEXP %in% c('k', 'K') ~ 1000,
            PROPDMGEXP %in% c('b', 'B') ~ 1000000000,
            PROPDMGEXP %in% c('h', 'H') ~ 100,
            PROPDMGEXP == '+' ~ 1,
            # Vector must be double because of limitations in `case_when`, so we
            # can't set a value to NA; use dummy code and convert later.
            TRUE ~ -999999
        ),
        crop_mult = case_when(
            PROPDMGEXP %in% c(0, 1, 2, 3, 4, 5, 6, 7, 8) ~ 10,
            CROPDMGEXP %in% c('m', 'M') ~ 1000000,
            CROPDMGEXP %in% c('k', 'K') ~ 1000,
            CROPDMGEXP %in% c('b', 'B') ~ 1000000000,
            CROPDMGEXP %in% c('h', 'H') ~ 100,
            CROPDMGEXP == '+' ~ 1,
            # Vector must be double because of limitations in `case_when`, so we
            # can't set a value to NA; use dummy code and convert later.
            TRUE ~ -999999
        ),
        # Change dummy value into NA
        across(ends_with("_mult"), ~na_if(., -999999)),
        crop_dmg = CROPDMG * crop_mult,
        prop_dmg = PROPDMG * prop_mult,
        tot_dmg = prop_dmg + crop_dmg
    )

# Switch back to dtplyr syntax and clean up
temp_data <- lazy_dt(temp_data)

# Filter out events which did not cause any injuries, fatalities or monetary
# damages
mdata <- temp_data %>% filter(
  (INJURIES > 0) | 
    (FATALITIES > 0) |
    (tot_dmg > 0)
) %>%
  # Apply data.table code
  as_tibble()

```
We can get the number of invalid `EXP` values resulting in `0`-value multipliers using the following code:

```{r invalid-exps}
# Property damage
inv_prop_exp <- filter(temp_data, PROPDMG != 0, prop_mult == 0) %>%
    select(PROPDMG, PROPDMGEXP, prop_mult, BGN_DATE) %>%
    nrow()

inv_crop_exp <- filter(temp_data, CROPDMG != 0, crop_mult == 0) %>%
    select(CROPDMG, CROPDMGEXP, crop_mult, BGN_DATE) %>%
    nrow()

# Clean up
rm(temp_data)
```

@@

skimr::skim(mdata$INJURIES)
cat("\n\n\n")
skimr::skim(mdata$FATALITIES)

@@

# Make a copy of data and cast as DT
dt_data <- as.data.table(lean_data)
# Define the number of rows that will be in the grouped DT. This is equal to the
# unique values of event, as we'll group by that.
denominator <- uniqueN(dt_data, "event")
# Apply final transformations to DT
sel_data <- dt_data[, .(
                        counter = .N, # Count elements in group
                        # Total monetary damages per group
                        total_dmg = sum(tot_dmg, na.rm = TRUE),
                        # Total fatalities
                        fatalities = sum(FATALITIES, na.rm = TRUE),
                        # Injuries
                        injuries = sum(INJURIES, na.rm = TRUE)
                        ),
                    # Grouping by event type
                    by = event][
                        ,
                        # Compute a "percent rank": rank events by number of
                        # occurrences (desc), then assign a percentile by
                        # dividing rank by denominator (i.e. nrows)
                        # Note double dot syntax to access calling scope
                        `:=`(perc_rank = frankv(counter)/..denominator,
                             # Same for monetary damage
                             dmg_rank = frankv(total_dmg)/..denominator,
                             # And life damage
                             inj_rank = frankv(injuries)/..denominator,
                             fat_rank = frankv(fatalities)/..denominator
                             )
                        ][
                            # Filter for frequency rank >=95th
                            perc_rank >= 0.90
                        ][
                            order(-counter)
                        ]

@@

sel_data %>% ggplot(
  aes(x = fct_reorder(event, perc_rank, .desc = T), y = counter)
) +
  geom_point() +
  scale_y_continuous(
    # limits = c(0, 100), 
    name = ("Frequency quantile\n")) +
  theme_bw()
